Climate-driven phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-07-21 agujournaltemplate.tex
In process- editing Andrews Comments

\begin{abstract}
Phytoplankton communities play a crucial role in ocean biogeochemical cycles. Climate drives changes in the physical environment that phytoplankton communities experience, thus impacting the productivity and diversity of marine ecosystems. Long-term time series of XXX can capture changes in climate and ecosystems, allowing us to understand the effects of environmental change on local communities. This study aims to describe and investigate the changes in the phytoplankton community of the Cariaco Basin, Venezuela, between 1995 and 2017. Water samples were taken during monthly cruises to determine plankton community composition and local water column conditions, which we integrated climate indice and meteorological data. 
Using clustering algorithms, we investigate the structure of the plankton community data, revealing two clusters that described a shift from high- to low-upwelling conditions in 2004, but also show a return to a community and conditions similar to those before the shift in 2014. However,  community diversity did not follow the same pattern, showing a decrease between 1998 and 2009. To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that a combination of in situ upwelling-related variables (sea surface temperature and nitrate concentration) and climate indices (AMO and MEI v.2) were the strongest predictors. Our study emphasizes the impact of long-term (decadal) climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and highlights the capability of ecosystems to recover from the effects of climate change.

@@

Microscopic algae drive... (Will write this Summary last, after draft is finalized)

@@

%Comments from Subhi:
    % - first 2 paragraphs, repetetive, put previously known parts into intro, just start from "the problem"
    % - streamline discussion, much shorter!
    % -

@@

A tropical coastal location that has been studied intensively in the last few decades is the Cariaco Basin, where a wide range of physical and biogeochemical parameters were consistently collected from 1995 to 2017 \cite{muller-karger_scientific_2019}. The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within one of the largest anoxic basins in the world oceans \cite{edgcomb_accessing_2011}. The Cariaco Basin is characterized by a marked seasonal cycle of upwelling, which is driven by easterly trade winds, with the highest levels of primary production observed from December through May when the Inter-Tropical Convergence Zone (ITCZ) moves southward. A smaller wind-driven secondary upwelling event occurs often in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}. During the upwelling period, the supply of deeper nutrient-rich waters supports high biological productivity, and the phytoplankton community is dominated by diatoms \cite{romero_seasonal_2009}, with dinoflagellates, coccolithophorids, and nanoplankton also contributing. In the low-wind, rainy season, the Cariaco Basin exhibits oligotrophic conditions, and the fraction of large phytoplankton cells is reduced. However, nano- and picophytoplankton cells are present in relatively high numbers throughout the year \cite{lorenzoni_characterization_2015}.

    % PARAGRAPH THREE - CHANGES IN CARIACO
    Since the inception of the time series in 1995, the ecosystem of the Cariaco Basin has undergone marked changes. \citeA{taylor_ecosystem_2012} analyzed the time series data up to 2010 and found  decadal trends of increasing sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors attributed to global warming. During the initial phase of the time series (1996–2004), the Cariaco Basin underwent consistent and pronounced seasonal upwelling cycles, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}. After 2004, a significant reduction in the strength of trade winds during the upwelling period \cite{taylor_ecosystem_2012} caused a regime change in the physical and biogeochemical system. This strongly affected the phytoplankton community and the entire ecosystem. Microscopy cell counts indicated a drastic reduction of large phytoplankton, particularly diatoms, as confirmed, albeit not with the same high significance, by analysis of pigment data \cite{pinckney_phytoplankton_2015}. In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, groups commonly found in stratified water columns \cite{lorenzoni_characterization_2015}. An analysis of pigment data indicated a deepening of the chlorophyll maximum, an increase in species diversity, and a reduction in the seasonal variability of community composition \cite{pinckney_phytoplankton_2015}.

    % PARAGRAPH FOUR - EXPLANATION OF CHANGES // OPEN QUESTIONS!!
    However, the shift in phytoplankton community composition associated with increased temperature and reduced upwelling intensities in year-start-year-end did not appear to be a permanent change in regime. Data from 2012 to 2017, the end phase of the CARIACO time series, indicated a return to strong upwelling intensities and to high numbers of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}. The impact of large-scale climatic processes on the observed shifts in the phytoplankton community remains unclear. The Atlantic Multidecadal Oscillation (AMO) has emerged as an important factor influencing the climate system in the region and the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}. The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}. The El Niño Southern Oscillation (ENSO) is teleconnected to the Caribbean Sea, given that during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures increase and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. However, using the time series available only until 2010, \citeA{taylor_ecosystem_2012} found no significant correlation between MEI v.2 and in situ observations, and did not test for correlations with the AMO index.

    % What drives the changes??? Bottom-up/ Top-down ... CLIMATE 
    \citeA{taylor_ecosystem_2012} cautioned that the observed changes cannot be clearly identified as "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation". Previous analyses of phytoplankton community diversity in the Cariaco Basin did not include time series data beyond 2011, and no detailed investigations have been undertaken yet on possible changes in diversity based on microscopy cell counts. Given these knowledge gaps, we aim to investigate 1) changes in the depth distribution of phytoplankton biomass and community diversity over the period from 1995 - 2017, 2) changes in the structure of the phytoplankton community data and potential connections with shifts observed in physical and biogeochemical variables, and 3) potential connections between long-term climatic oscillations (ENSO MEI v.2 and AMO) and changes in the phytoplankton community. Our analysis focuses exclusively on bottom-up drivers of community changes, because these data are available throughout the entire time series and at a monthly resolution. Our study builds on and expands substantially the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017. Understanding the regional dynamics of the Cariaco Basin is crucial for unveiling the ecological changes affecting this tropical marine ecosystem and the variables driving these changes.

\section{Materials and Methods}
% Here is text on Materials and Methods.
%
\subsection{Study Site}
    %Talk about the Basin shortly, then about the time series
    The Cariaco Ocean Time Series Program was established to explore the relationship between ocean surface processes and the sinking flux of particulate carbon. The Cariaco Basin (Figure~\ref{fig:map}) is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin, situated on the northeastern Venezuelan continental shelf, with a sill to the north, running almost parallel to the coastline at a mean depth of around \qty{100}{m}. This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at the BATS and HOT sites, resulting in the formation of anoxic conditions below \qty{\sim 275}{m} depth.

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/Map_CAR.pdf}
\caption{Bathymetric map of the CARIACO Basin in the Caribbean Ocean. The station is located at \ang{10.50}N, \ang{64.67}W). A sill, running almost parallel to the coast at an average depth of \qty{100}{m}, from the Isla de Magarita and the Isla la Tortuga and beyond, encloses the Cariaco basin from the North.}
\label{fig:map}
\end{figure}

\subsection{In-situ biological and environmental data}
    <Add a topic sentence stating what in situ data are use for.> Samples were collected as part of the Cariaco Ocean Time Series Program on monthly research cruises at a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.66}W) \cite{muller-karger_scientific_2019}.
    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, along with other deployments and samples not utilized in this work. Details about the sampling procedures, analytical methods, data quality control, and inter-calibration procedures have already been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.

    All samples were collected using a 12-bottle rosette system equipped with a CTD (SeaBird, model SBE 25) \cite{astor_yrene_m_handbook_2013}. Water samples for chlorophyll-a and nutrient analysis were taken monthly during the first cast of each cruise. These samples were collected at eight standard depths within the upper 100 meters (1, 7, 15, 25, 35, 55, 75, and 100 m).

@@

    For the identification of phytoplankton species by microscopy, \qty{500}{\milli\liter} seawater samples were collected at the same depths in HDPE bottles and preserved with a \qty{5}{\%} (final concentration) formalin solution neutralized with sodium tetraborate. Species counts were performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using \qty{100}{\milli\liter} sedimentation chambers with a 48-hour settling period. Phytoplankton identification and enumeration by microscopy were restricted to microphytoplankton (cells $>$ \qty{20}{\micro \meter}) \cite{mutshinda_environmental_2013}.
    %from Mutshinda et al. 2013
    Quality control of the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level (with the exception of nanoflagellate cell counts), correcting spelling variations, and grouping synonyms using taxonomic information from the World Register of Marine Species (\url{marinespecies.org}). 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera.

    Data collected with Niskin bottles, CTD, and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047}. 
    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (no more than three months) in coverage. In total, we used a complete set of 254 monthly data points from YearX-YearY, for which all variables were available.

\subsection{Meteorological and climate data}
    %from Taylor et al. 2012
    In addition to oceanographic environmental data, we also examine meteorological and climatic data that are known to influence ocean conditions and marine ecosystems. Zonal wind stress (u, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters  \cite{rueda-roa_southern_2013}. We extracted the u wind component at \qty{10}{\meter} from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}. For simplicity, we refer to this variable as "wind speed" throughout the article. The data for precipitation and evaporation are also extracted from the ERA5 dataset for the aforementioned grid cell. Evaporation provides an estimate of the accumulated amount of water that has evaporated from Earth's surface, while precipitation is an estimate for total precipitation, comprising all water that falls on Earth's surface as rain or snow.

    In addition the wind data, we examined two climatic indices, each of which provide information on the basin-scale climatic conditions and consequently the broad-scale forcing of marine ecosystems. The monthly Atlantic Multidecadal Oscillation (AMO) index data (10-year low-pass), which represent area-averaged, detrended low-pass filtered North Atlantic SST anomalies over what area, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo}.

    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure, SLP); sea surface temperature, SST; zonal and meridional components of surface wind; and outgoing long-wave radiation, OLR)) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W), were obtained from \url{https://psl.noaa.gov/enso/mei/}. In our analysis, bi-monthly values were aligned with the tail month (December-January is treated as the value for January).

@@

\subsection{Data analysis}    
    To provide an overview of the available data and the general trends, we calculated annual means for relevant variables and normalized the data range as z-scores ($z = \frac{(x – \mu)}{\sigma}$). Cruises from 1995, 2016, and 2017 were excluded because insufficient coverage in these years prevented us from producing reliable yearly mean estimates.

    We calculated three metrics of phytoplankton community diversity. First, we calculated genus richness, which reduces taxonomic resolution cut creates a more homogeneous metric that is robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008, de2020higher}. Second and third, we calculated the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou's evenness $J' = -\sum_i p_i ln( p_i )/ln(H')$, respectively, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling.

@@

    \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
    To quantify the effects of bottom-up drivers in the Cariaco Basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis. Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables (here: monthly observed genus counts) to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then it averages these to determine the overall importance of each predictor. The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. This analysis was conducted using monthly measurements integrated over a depth of 100 m for observed genus counts, ensuring the inclusion of all available data. The drivers/predictor variables considered in the analysis were the AMO and MEI v.2 indices, wind speed, evaporation, precipitation, \qty{21}{\celsius} isotherm, SST, salinity, and nutrient concentrations ($\mathrm{[NO]}_3$, $\mathrm{[PO]}_4$ and $\mathrm{[SiO]}_4$). 
    The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology accounts for by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}. 
    We show a model run without any time lags in  supplementary material (Figure \ref{fig:sup:GFoutput_nolags}). To consider the temporal lag in effect, particularly of climate indices, we included time lags of up to 6 months for all climate variables, as well as 3-month lags for in-situ measurements in exploratory runs. We chose time lags that scored highest in the goodness-of-fit statistics for the final model run (Table \ref{tab:sup:GFlagtests} in Appendix).

@@

Create subsections in results. This one could be "Interannual changes in the environment and ecosystem"
% Figure 2
Coordinated, interannual changes occur in normalized yearly means of all environmental variables considered, the logarithmic transformation of chlorophyll \textit{a}, and phytoplankton cell counts per functional group (Figure \ref{fig:zscore}). The patterns emerging from wind speed, evaporation, precipitation, SST, and \qty{21}{\celsius} isotherm depth clearly show the years of positive anomalies of wind (blue) and negative anomalies for SST (red) and upwelling (red, i.e., shallower depth of the isotherm). These patterns are reflected in the mode of the AMO, where a positive anomaly correlates with weak upwelling regimes (see also Figure \ref{fig:sup:correlation} in Appendix). The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series but show differing inter-annual oscillations. Annual mean anomalies of nitrate and phosphate appear to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some variability in mean phosphate concentration. Silicate does not follow a clear pattern but shows an elevated mean concentration between 2008 and 2013, which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion due to reduced abundances of diatoms.  <I think the most obvious pattern in Fig. 2 has not yet been discussed: Chl and phyto are high in 1996-2004, then low in 2004-2013, followed by a switch in 2014-2015. Diversity is high when chl/phyto are low. I think you'll want to describe these patterns as they are interesting.>

>>>>

\begin{abstract}
Phytoplankton communities play a crucial role in ocean biogeochemical cycles. Climate drives changes in the physical environment that phytoplankton communities experience, thus impacting the productivity and diversity of marine ecosystems. Long-term time series of phytoplankton can capture changes in climate and ecosystems, allowing us to understand the effects of environmental change on local communities. This study aims to describe and investigate the changes in the phytoplankton community of the Cariaco Basin, Venezuela, between 1995 and 2017. Water samples were taken during monthly cruises to determine plankton community composition and local water column conditions, which we integrated with climate indices and meteorological data. 
Using clustering algorithms, we investigate the structure of the phytoplankton community data, revealing two clusters that described a shift from high- to low-upwelling conditions in 2004, but also show a return to a community and conditions similar to those before the shift in 2014. However,  community diversity did not follow the same pattern, showing a decrease between 1998 and 2009. To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that a combination of in situ upwelling-related variables (sea surface temperature and nitrate concentration) and climate indices (AMO and MEI v.2) were the strongest predictors. Our study emphasizes the impact of long-term (decadal) climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and highlights the capability of ecosystems to recover from the effects of climate change.

@@

% NO MORE THAN 200 Words
% Topic Overview
Phytoplankton are microscopic algae that form the foundation of marine food webs and ecosystems. Their growth and community composition are impacted by the local physical and chemical environment, which is determined by currents and climate, for example, through processes such as wind-driven upwelling of nutrients from the deeper ocean. Time series data of phytoplankton biomass and community composition can be very helpful in untangling the link between global and local climate and its effects on the phytoplankton community.
% Paper Overview
In this study, we set out to analyze changes in the phytoplankton community between 1995 and 2017 at a time series location in a large coastal basin in Venezuela, the Cariaco Basin. Previous analysis of the data found strong shifts within the phytoplankton community that could be related to impacts of global warming. Our main question in approaching the data was whether we can confirm this shift in community composition or whether other climate processes can explain the observed patterns. 
% Findings Summary
We found that the phytoplankton community could be separated into two groups. The first group includes an early and late period (1995-2003 and 2014-2017) that was characterized by robust phytoplankton growth driven by relatively strong upwelling of nutrients. The second group is the intermediate period, where a reduction in upwelling and phytoplankton biomass was observed. Additionally, we related physical and climatic variables, such as sea surface temperature and climate indices that relate to long-term oscillations in the North Atlantic, to the changes in community composition. By allowing for a time lag, we found that the Atlantic Meridional Oscillation (AMO) index was the strongest predictor of community changes, with temperature and 
% Key Takeaways

@@

@@

A tropical coastal location that has been studied intensively in the last few decades is the Cariaco Basin, where a wide range of physical and biogeochemical parameters were consistently collected from 1995 to 2017 \cite{muller-karger_scientific_2019}. The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within one of the largest anoxic basins in the world's oceans \cite{edgcomb_accessing_2011}. The Cariaco Basin is characterized by a marked seasonal cycle of upwelling, which is driven by easterly trade winds, with the highest levels of primary production observed from December through May when the Inter-Tropical Convergence Zone (ITCZ) moves southward. A smaller wind-driven secondary upwelling event occurs often in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}. During the upwelling period, the supply of deeper nutrient-rich waters supports high biological productivity, and the phytoplankton community is dominated by diatoms \cite{romero_seasonal_2009}, with dinoflagellates, coccolithophorids, and nanoplankton also contributing. In the low-wind, rainy season, the Cariaco Basin exhibits oligotrophic conditions, and the fraction of large phytoplankton cells is reduced. However, nano- and picophytoplankton cells are present in relatively high numbers throughout the year \cite{lorenzoni_characterization_2015}.

% PARAGRAPH THREE - CHANGES IN CARIACO
Since the inception of the time series in 1995, the ecosystem of the Cariaco Basin has undergone marked changes. \citeA{taylor_ecosystem_2012} analyzed the time series data up to 2010 and found decadal trends of increasing sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors attributed to global warming. During the initial phase of the time series (1996–2004), the Cariaco Basin underwent consistent and pronounced seasonal upwelling cycles, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}. After 2004, a significant reduction in the strength of trade winds during the upwelling period \cite{taylor_ecosystem_2012} caused a regime change in the physical and biogeochemical system. This strongly affected the phytoplankton community and the entire ecosystem. Microscopy cell counts indicated a drastic reduction of large phytoplankton, particularly diatoms, as confirmed, albeit not with the same high significance, by analysis of pigment data \cite{pinckney_phytoplankton_2015}. In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, groups commonly found in stratified water columns \cite{lorenzoni_characterization_2015}. An analysis of pigment data indicated a deepening of the chlorophyll maximum, an increase in pigment diversity, and a reduction in the seasonal variability of community composition \cite{pinckney_phytoplankton_2015}.

% PARAGRAPH FOUR - EXPLANATION OF CHANGES // OPEN QUESTIONS!!
However, the shift in phytoplankton community composition associated with increased temperature and reduced upwelling intensities from 2004 to 2013 did not appear to be a permanent change in regime. Data from 2013 to 2017, the end phase of the CARIACO time series, indicated a return to strong upwelling intensities and high numbers of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}. The impact of large-scale climatic processes on the observed shifts in the phytoplankton community remains unclear. The Atlantic Multidecadal Oscillation (AMO) has emerged as an important factor influencing the climate system in the region and the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}. The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}. The El Niño Southern Oscillation (ENSO) is teleconnected to the Caribbean Sea, given that during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures increase and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. However, using the time series available only until 2010, \citeA{taylor_ecosystem_2012} found no significant correlation between MEI v.2 and in situ observations, and did not test for correlations with the AMO index.

% What drives the changes??? Bottom-up/ Top-down ... CLIMATE 
\citeA{taylor_ecosystem_2012} cautioned that the observed changes cannot be clearly identified as "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation". Previous analyses of phytoplankton community diversity in the Cariaco Basin did not include time series data beyond 2011, and no detailed investigations have yet been undertaken on possible changes in diversity based on microscopy cell counts. Given these knowledge gaps, we aim to investigate 1) changes in the depth distribution of phytoplankton biomass and community diversity over the period from 1995 to 2017, 2) changes in the structure of the phytoplankton community data and potential connections with shifts observed in physical and biogeochemical variables, and 3) potential connections between long-term climatic oscillations (ENSO MEI v.2 and AMO) and changes in the phytoplankton community. Our analysis focuses exclusively on bottom-up drivers of community changes because these data are available throughout the entire time series and at a monthly resolution. Our study builds on and substantially expands the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017. Understanding the regional dynamics of the Cariaco Basin is crucial for unveiling the ecological changes affecting this tropical marine ecosystem and the variables driving these changes.

\section{Materials and Methods}
% Here is text on Materials and Methods.
%
\subsection{Study Site}
    %Talk about the Basin shortly, then about the time series
    The Cariaco Ocean Time Series Program was established to explore the relationship between ocean surface processes and the sinking flux of particulate carbon. The Cariaco Basin (Figure~\ref{fig:map}) is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin situated on the northeastern Venezuelan continental shelf, with a sill to the north running almost parallel to the coastline at a mean depth of around \qty{100}{m}. This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at the BATS and HOT sites, resulting in the formation of anoxic conditions below \qty{\sim 275}{m} depth.

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/Map_CAR.pdf}
\caption{Bathymetric map of the CARIACO Basin in the Caribbean Ocean. The station is located at \ang{10.50}N, \ang{64.67}W).}
\label{fig:map}
\end{figure}

\subsection{In-situ biological and environmental data}
    The CARIACO Ocean Time Series Program collected biological and environmental data, which we utilized here to examine connections between environmental and ecological change in the region. Samples were collected on monthly research cruises at a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.67}W) \cite{muller-karger_scientific_2019}.
    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, along with other deployments and samples not utilized in this work. Details about the sampling procedures, analytical methods, data quality control, and inter-calibration procedures have already been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.

    All samples were collected using a 12-bottle rosette system equipped with a CTD (SeaBird, model SBE 25) \cite{astor_yrene_m_handbook_2013}. Water samples for chlorophyll \textit{a} and nutrient analysis were taken monthly during the first cast of each cruise. These samples were collected at eight standard depths within the upper 100 meters (1, 7, 15, 25, 35, 55, 75, and 100 m).

@@

    For the identification of phytoplankton species by microscopy, \qty{500}{\milli\liter} seawater samples were collected at depths of 1, 7, 15, 25, 35, 55, 75, and 100 m in HDPE bottles and preserved with a \qty{5}{\%} (final concentration) formalin solution neutralized with sodium tetraborate. Species counts were performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using \qty{100}{\milli\liter} sedimentation chambers with a 48-hour settling period. Phytoplankton identification and enumeration by microscopy were restricted to microphytoplankton (cells $>$ \qty{20}{\micro \meter}) \cite{mutshinda_environmental_2013}.
    %from Mutshinda et al. 2013
    Quality control of the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level (with the exception of nanoflagellate cell counts), correcting spelling variations, and grouping synonyms using taxonomic information from the World Register of Marine Species (\url{marinespecies.org}). A total of 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera.

    All in-situ data, inlcuding Niskin bottle, CTD, and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047}. 
    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (no more than three months) in coverage. In total, we used a complete set of 254 monthly data points from 1995 to 2017, for which all variables were available.

\subsection{Meteorological and climate data}
    %from Taylor et al. 2012
    In addition to oceanographic environmental data, we also examine meteorological and climatic data that are known to influence ocean conditions and marine ecosystems. Zonal wind (\textit{u}, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters \cite{rueda-roa_southern_2013}. We extracted the variable u10, the \qty{10}{\meter} eastward component of wind speed from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}. For simplicity, we refer to this variable as "wind speed" throughout the article. The data for precipitation and evaporation are also extracted from the ERA5 dataset for the aforementioned grid cell. Evaporation provides an estimate of the accumulated amount of water that has evaporated from Earth's surface, while precipitation is an estimate of total precipitation, comprising all water that falls on Earth's surface as rain or snow.

    In addition to the wind data, we examined two climatic indices, each of which provides information on the basin-scale climatic conditions and, consequently, the broad-scale forcing of marine ecosystems. The monthly Atlantic Multidecadal Oscillation (AMO) index data (10-year low-pass), which represent area-averaged, detrended low-pass filtered North Atlantic SST anomalies (\ang{0}N-\ang{60}N), were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo}.

    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure, SLP; sea surface temperature, SST; zonal and meridional components of surface wind; and outgoing long-wave radiation, OLR) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W), were obtained from \url{https://psl.noaa.gov/enso/mei/}. In our analysis, bi-monthly values were aligned with the tail month (December-January is treated as the value for January).

@@

\subsection{Data analysis}    
    To provide an overview of the available data and highlight the interannual trends and changes in the time series data, we calculated annual means for relevant variables and normalized the data range as z-scores ($z = \frac{(x – \mu)}{\sigma}$). Cruises from 1995, 2016, and 2017 were excluded because insufficient coverage (i.e. more than six missing monthly samples) in these years prevented us from producing reliable yearly mean estimates.

    We calculated three metrics of phytoplankton community diversity. First, we calculated genus richness, which reduces taxonomic resolution but creates a more homogeneous metric that is robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008, de2020higher}. Second and third, we calculated the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou's evenness $J' = -\sum_i p_i ln( p_i )/ln(H')$, respectively, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling.

@@

    \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
    To quantify the effects of bottom-up drivers in the Cariaco Basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis. Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables (here: monthly observed genus counts) to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor. The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. This analysis was conducted using monthly measurements integrated over a depth of 100 m for observed genus counts, ensuring the inclusion of all available data. The drivers/predictor variables considered in the analysis were the AMO and MEI v.2 indices, wind speed, evaporation, precipitation, \qty{21}{\celsius} isotherm, SST, salinity, and nutrient concentrations ($\mathrm{[NO]}_3$, $\mathrm{[PO]}_4$ and $\mathrm{[SiO]}_4$). 
    The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology accounts for by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}. 
    We show a model run without any time lags in  supplementary material (Figure \ref{fig:sup:GFoutput_nolags}). To consider the temporal lag in effect, particularly of climate indices, we included time lags of up to 6 months for all climate variables, as well as 3-month lags for in-situ measurements in exploratory runs. We chose time lags that scored highest in the goodness-of-fit statistics for the final model run (Table \ref{tab:sup:GFlagtests} in Appendix).

@@

Create subsections in results. This one could be "Interannual changes in the environment and ecosystem"
% Figure 2
Coordinated, interannual changes occur in normalized yearly means of all environmental variables considered, the logarithmic transformation of chlorophyll \textit{a}, and phytoplankton cell counts per functional group (Figure \ref{fig:zscore}). The patterns emerging from wind speed, evaporation, precipitation, SST, and \qty{21}{\celsius} isotherm depth clearly show the years of positive anomalies of wind (blue) and negative anomalies for SST (red) and upwelling (red, i.e., shallower depth of the isotherm). These patterns are reflected in the mode of the AMO, where a positive anomaly correlates with weak upwelling regimes (see also Figure \ref{fig:sup:correlation} in Appendix). The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series but show differing interannual oscillations. Annual mean anomalies of nitrate and phosphate appear to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some variability in mean phosphate concentration. Silicate does not follow a clear pattern but shows an elevated mean concentration between 2008 and 2013, which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion due to reduced abundances of diatoms.  <I think the most obvious pattern in Fig. 2 has not yet been discussed: Chl and phyto are high in 1996-2004, then low in 2004-2013, followed by a switch in 2014-2015. Diversity is high when chl/phyto are low. I think you'll want to describe these patterns as they are interesting.>
